Location and Scale-Invariant Power Transformations for Transforming Data to Normality

### 2024-08-23 manuscript/manuscript.Rmd
Changes to manuscript.

# Methods

@@

```{=latex}
\begin{equation}
\begin{split}
\mathcal{l}_{\text{rBC}}^{\lambda, x_0, s} = & -\frac{1}{2} \left(\sum_{i=1}^n w_i \right) \log \left[ 2 \pi \sigma_w^2 \right] -\frac{1}{2 \sigma_w^2} \sum_{i=1}^n w_i \left( \phi^{\lambda, x_0, s}(x_i) - \mu_w \right)^2 \\
& - \lambda \left( \sum_{i=1}^n w_i \right) \log s + \left( \lambda - 1 \right) \sum_{i=1}^n w_i \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}
```
where $\mu_w$ and $\sigma^2_w$ are the weighted mean and weighted variance of the Box-Cox transformed feature $\phi_{\text{BC}}^{\lambda, x_0, s} (\mathbf{X})$:

@@

```{=latex}
\begin{equation}
\begin{split}
\mathcal{l}_{\text{rYJ}}^{\lambda, x_0, s} = & -\frac{1}{2} \left(\sum_{i=1}^n w_i \right) \log \left[ 2 \pi \sigma_w^2 \right] -\frac{1}{2 \sigma_w^2} \sum_{i=1}^n w_i \left( \phi^{\lambda, x_0, s}(x_i) - \mu_w \right)^2 \\
& - \left( \sum_{i=1}^n w_i \right) \log s + (\lambda - 1) \sum_{i=1}^n w_i \sgn(x_i - x_0) \log \left[1 + \frac{|x_i - x_0|}{s} \right]
\end{split}
\end{equation}
```
where $\mu_w$ and $\sigma^2_w$ are the weighted mean and weighted variance of the Yeo-Johnson transformed feature $\phi_{\text{YJ}}^{\lambda, x_0, s} (\mathbf{X})$:

@@

In this section we simulate data to assess invariance to location and scale of the proposed power transformations, weighting for robust transformations, and to develop the empirical central normality test.
The $\lambda$ parameter for conventional power transformations (Eqn. \ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}), as well as $\lambda$, $x_0$ and $s$ parameters for location and scale invariant power transformations (Eqn. \ref{eqn:box-cox-invariant} and \ref{eqn:yeo-johnson-invariant}) are estimated using the BOBYQA algorithm for derivative-free bound constraint optimisation [@Powell2009-zb] through maximum likelihood estimation.

@@

To assess whether the proposed power transformations lead to values of $\lambda$ that are invariant to location and scale of the distribution, we simulate three different distributions. The first distribution is a normal distribution, as previously shown in Figure \ref{fig:decreased-normality}.
We first randomly draw $10000$ values from a normal distribution: $\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{N}\left(0, 1\right)$, or equivalently $\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.5, 2\right)$.
The second distribution is a right-skewed normal distribution $\mathbf{X}_{\text{right}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.2, 2\right)$.
The third distribution is a left-skewed normal distribution $\mathbf{X}_{\text{left}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.8, 2\right)$.
We then compute transformation parameter $\hat{\lambda}$ using the original definitions (equations \ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}) and the location and scale invariant definitions (equations \ref{eqn:box-cox-invariant} and \ref{eqn:yeo-johnson-invariant}) for each distribution.
To assess location invariance, a positive value $d_{\text{shift}}$ is added to each distribution with $d_{\text{shift}} \in [1, 10^6]$. Similarly, to assess scale invariance, each distribution is multiplied by a positive value $d_{\text{scale}}$, where $d_{\text{scale}} \in [1, 10^6]$.

@@

The result is shown in Figure \ref{fig:shifted-distributions}. For each distribution, transformation parameter $\hat{\lambda}$ varies with $d_{\text{shift}}$ and $d_{\text{scale}}$ when estimated using the original definitions.
In contrast, estimation of $\lambda$ using invariant power transformations was invariant to $d_{\text{shift}}$ and $d_{\text{scale}}$. 
While the effect of location on the estimated value of $\lambda$ using conventional power transformations is noticeable, scale also affects $\lambda$ for features that are not normally distributed.
Figure \ref{fig:shifted-distributions} also shows that estimated $\lambda$ for Box-Cox power transformations in left-skewed distributions diverges from that found for invariant Box-Cox power transformations.

## Empirical distribution yields consistent robust transformations

Outliers may be presented in the data, and affect estimation of power transformation parameters.
The log-likelihood function is weighted to assign less weight to outlier instances.
Weighting is based on either probabilities of the empirical distribution of the original feature, the z-score of the transformed feature values, or the residual error between the z-score of the transformed feature values and their expected z-score based on the normal distribution.
A weight is then assigned to each instance using one of three weighting functions: step, triangle and tapered cosine.
The weighting functions have one (step) or two (triangle, tapered cosine) parameters.
Here we investigate how these parameters should be set to find transformation parameter $\hat{\lambda}^r$ that are robust to outliers.

Let us randomly draw $m_d=100$ asymmetric generalised normal distributions.
Each distribution is parametrised with a randomly chosen skewness parameter $\alpha \sim U\left(0.01, 0.99\right)$ and shape parameter $\beta \sim U\left(1.00, 5.00 \right)$.
Location and scale parameters are set as $\mu = 0$ and $\sigma = 1$, respectively.
$n = \lceil 10^\gamma \rceil$ values are then randomly drawn, with $\gamma \sim U\left(2, 4\right)$, i.e., between $100$ and $10000$ values are drawn to create $\mathbf{X}_i$.

Outlier values are then drawn to randomly replace a fraction of the values of $\mathbf{X}_i$.
This is repeated $m_{out} = 10$ times, with outlier fractions regularly spaced in $[0.00, 0.10]$.
Thus up to 10 percent of the values may be replaced by outliers.
Outlier values are set according to @Tukey1977-xm.
Let $x^{*} \sim U\left(-2, 2\right)$.
Then the corresponding outlier value is:

@@

Here, $Q_1$, $Q_3$ and $\text{IQR}$ are the first quartile, third quartile and interquartile range of $\mathbf{X}_i$, respectively.
Outlier values randomly replace values in $\mathbf{X}_i$ to create $\mathbf{X}_{i,j}$.

To find the optimal values for the weighting function parameters $k_1$ and $k_2$ (if applicable), we minimise the absolute difference between the $\hat{\lambda}^r$ parameter obtained for transformation in the presence of outliers, and the $\hat{\lambda}_0$ parameter obtained using the non-robust modified transformation in absence of outliers:

```{=latex}
\begin{equation}
\left\{ \hat{k}_1, \hat{k}_2 \right\} = \argmin_{k_1, k_2} \sum_{i=1}^{m_d} \sum_{j=1}^{m_{out}} \left| \hat{\lambda}^r \left(\mathbf{X}_{i, j}; k_1, k_2 \right) - \hat{\lambda}_{0} \left(\mathbf{X}_i \right) \right|
\end{equation}
```
The double sum in right-hand side of the above equation can be considered as the loss that should be minimised.
As mentioned earlier, transformation parameters are estimated using the BOBYQA algorithm for derivative-free bound constraint optimisation [@Powell2009-zb] through maximum likelihood estimation.

The optimal weighting function parameters for weighted MLE are shown in Tables \ref{tab:optimal-weighting-parameters-box-cox} and \ref{tab:optimal-weighting-parameters-yeo-johnson} for robust, location- and scale-invariant Box-Cox and Yeo-Johnson transformations, respectively.

@@

Figure \ref{fig:optimised-weighting-function-parameters} shows the distribution of errors $\left| \hat{\lambda}^r - \hat{\lambda}_0 \right|$ for non-robust transformations, as well as robust transformations using the optimal weighting function parameters. In the presence of outliers on either side of the distribution, the Yeo-Johnson transformation yields smaller errors than the Box-Cox transformation. For both transformations, weighting based on empirical probabilities yields the most consistent $\lambda$ parameter estimates in the presence of outliers. Other methods did not outperform the non-robust power transformation method.

>>>>

# Theory

In this section, we will first introduce location and scale invariant versions of the Box-Cox and Yeo-Johnson transformations.
Subsequently, we define weighted location and scale invariant transformations and weighting methods that allow for defining robust transformations.
For the purpose of simulation we then define the quantile function for asymmetric generalised normal distributions.
Finally, we define the overall framework for the empirical central normality test.

@@

```{=latex}
\begin{equation}
\label{eqn:box-cox-weighted-invariant-log-likelihood}
\begin{split}
\mathcal{l}_{\text{rBC}}^{\lambda, x_0, s} = & -\frac{1}{2} \left(\sum_{i=1}^n w_i \right) \log \left[ 2 \pi \sigma_w^2 \right] -\frac{1}{2 \sigma_w^2} \sum_{i=1}^n w_i \left( \phi^{\lambda, x_0, s}(x_i) - \mu_w \right)^2 \\
& - \lambda \left( \sum_{i=1}^n w_i \right) \log s + \left( \lambda - 1 \right) \sum_{i=1}^n w_i \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}
```
where $\mu_w$ and $\sigma^2_w$ are the weighted mean and weighted variance of the Box-Cox transformed feature $\phi_{\text{BC}}^{\lambda, x_0, s} (\mathbf{X})$:

@@

```{=latex}
\begin{equation}
\label{eqn:yeo-johnson-weighted-invariant-log-likelihood}
\begin{split}
\mathcal{l}_{\text{rYJ}}^{\lambda, x_0, s} = & -\frac{1}{2} \left(\sum_{i=1}^n w_i \right) \log \left[ 2 \pi \sigma_w^2 \right] -\frac{1}{2 \sigma_w^2} \sum_{i=1}^n w_i \left( \phi^{\lambda, x_0, s}(x_i) - \mu_w \right)^2 \\
& - \left( \sum_{i=1}^n w_i \right) \log s + (\lambda - 1) \sum_{i=1}^n w_i \sgn(x_i - x_0) \log \left[1 + \frac{|x_i - x_0|}{s} \right]
\end{split}
\end{equation}
```
where $\mu_w$ and $\sigma^2_w$ are the weighted mean and weighted variance of the Yeo-Johnson transformed feature $\phi_{\text{YJ}}^{\lambda, x_0, s} (\mathbf{X})$:

@@

In this section we used simulated data to assess invariance to location and scale of the proposed power transformations, weighting for robust transformations, and to develop the empirical central normality test.
The $\lambda$ parameter for conventional power transformations (Eqn. \ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}), as well as $\lambda$, $x_0$ and $s$ parameters for location and scale invariant power transformations (Eqn. \ref{eqn:box-cox-invariant} and \ref{eqn:yeo-johnson-invariant}) were estimated using the BOBYQA algorithm for derivative-free bound constraint optimisation [@Powell2009-zb] through maximum likelihood estimation.

@@

To assess whether the proposed power transformations lead to values of $\lambda$ that are invariant to location and scale of the distribution, we simulated three different distributions. The first distribution was a normal distribution, as previously shown in Figure \ref{fig:decreased-normality}.
We first randomly drew $10000$ values from a normal distribution: $\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{N}\left(0, 1\right)$, or equivalently $\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.5, 2\right)$.
The second distribution was a right-skewed normal distribution $\mathbf{X}_{\text{right}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.2, 2\right)$.
The third distribution was a left-skewed normal distribution $\mathbf{X}_{\text{left}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.8, 2\right)$.
We then computed transformation parameter $\lambda$ using the original definitions (equations \ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}) and the location and scale invariant definitions (equations \ref{eqn:box-cox-invariant} and \ref{eqn:yeo-johnson-invariant}) for each distribution.
To assess location invariance, a positive value $d_{\text{shift}}$ was added to each distribution with $d_{\text{shift}} \in [1, 10^6]$.
Similarly, to assess scale invariance, each distribution was multiplied by a positive value $d_{\text{scale}}$, where $d_{\text{scale}} \in [1, 10^6]$.

@@

The result is shown in Figure \ref{fig:shifted-distributions}.
For each distribution, transformation parameter $\lambda$ varied with $d_{\text{shift}}$ and $d_{\text{scale}}$ when estimated using the original definitions.
The effect of location on the estimated $\lambda$ value for conventional power transformations is noticeable for all three distributions.
However scale only affected $\lambda$ for features that were sampled from a skewed distribution.
In contrast, estimation of $\lambda$ using invariant power transformations was invariant to both $d_{\text{shift}}$ and $d_{\text{scale}}$.

## Robust transformations

Outliers may be present in data and affect estimation of power transformation parameters.
The log-likelihood function can be weighted to assign less weight to outlier instances, see equations \ref{eqn:box-cox-weighted-invariant-log-likelihood} and \ref{eqn:yeo-johnson-weighted-invariant-log-likelihood}.
We proposed three weighting methods: probabilities of the empirical distribution of the original feature, the z-score of the transformed feature values, or the residual error between the z-score of the transformed feature values and their expected z-score based on the normal distribution.
The weight itself is then assigned to each instance using one of three weighting functions: step, triangle and tapered cosine.
The weighting functions have one (step) or two (triangle, tapered cosine) parameters.
Here we investigate how these parameters should be set to find robust transformation parameter $\lambda_r$.

We randomly drew $m_d=100$ asymmetric generalised normal distributions.
Each distribution was parametrised with a randomly chosen skewness parameter $\alpha \sim U\left(0.01, 0.99\right)$ and shape parameter $\beta \sim U\left(1.00, 5.00 \right)$.
Location and scale parameters were set as $\mu = 0$ and $\sigma = 1$, respectively.
$n = \lceil 10^\gamma \rceil$ instances were then randomly drawn, with $\gamma \sim U\left(2, 4\right)$, i.e., between $100$ and $10000$ values are drawn to create $\mathbf{X}_i$.

Outlier values were then drawn to randomly replace a fraction of the values of $\mathbf{X}_i$.
This was repeated $m_{out} = 10$ times, with outlier fractions regularly spaced in $[0.00, 0.10]$.
Thus up to 10 percent of the values could be replaced by outliers.
Outlier values were set according to @Tukey1977-xm, as follows.
Let $x^{*} \sim U\left(-2, 2\right)$.
Then the corresponding outlier value was:

@@

$Q_1$, $Q_3$ and $\text{IQR}$ are the first quartile, third quartile and interquartile range of $\mathbf{X}_i$, respectively.
Outlier values randomly replaced values in $\mathbf{X}_i$ to create $\mathbf{X}_{i,j}$.

To find the optimal values for the weighting function parameters $k_1$ and $k_2$ (if applicable), we minimised the absolute difference between the $\lambda_{r}$ parameter obtained for transformation in the presence of outliers, and the $\lambda_0$ parameter obtained using the non-robust transformations in absence of outliers:

```{=latex}
\begin{equation}
\label{eqn:minimisation-weighting}
\left\{ \hat{k}_1, \hat{k}_2 \right\} = \argmin_{k_1, k_2} \sum_{i=1}^{m_d} \sum_{j=1}^{m_{out}} \left| \lambda_r \left(\mathbf{X}_{i, j}; k_1, k_2 \right) - \lambda_0 \left(\mathbf{X}_i \right) \right|
\end{equation}
```

Minimisation was conducted using the BOBYQA algorithm for derivative-free bound constraint optimisation [@Powell2009-zb].
The resulting weighting function parameters for weighted MLE are shown in Tables \ref{tab:optimal-weighting-parameters-box-cox} and \ref{tab:optimal-weighting-parameters-yeo-johnson} for robust location and scale invariant Box-Cox and Yeo-Johnson transformations, respectively.

@@

Figure \ref{fig:optimised-weighting-function-parameters} shows the distribution of errors $\left| \lambda_r - \lambda_0 \right|$ for non-robust and robust transformations using the optimal weighting function parameters.
In the presence of outliers, the Yeo-Johnson transformation yielded smaller errors than the Box-Cox transformation.
For both transformations, weighting based on empirical probabilities yielded the most consistent $\lambda$ parameter estimates in the presence of outliers.
Other methods did not outperform the non-robust transformation method.
